Classifying galaxy spectra at $0.5<z<1$: an unsupervised approach

### 2016-10-26 sections0.07/intro0.07.tex
Final corrections after referee report.

%%Creating templates and NIR_UV templates
Several attempts have been made to create a complete set of templates for categorizing the spectral type of galaxies using \textbf{\textcolor{red}{data from}} nearby galaxies~(e.g.~\citealt{Kinney93}, \citealt[][hereafter \citetalias{Kinney96}]{Kinney96}, \citealt{Bershady00}, \citealt{Mannucci01}). 
Based on their usage, these templates are restricted to certain wavelengths.
In near-infrared (NIR) to ultraviolet (UV) wavelengths \textbf{\textcolor{red}{(the region where the energy output of stars peaks)}}, the SED contains information about the main physical properties of galaxies, e.g., age, star formation rate (SFR), stellar mass, metallicity, and interstellar medium.

@@

Advances in techniques and detectors have resulted in more detailed \textbf{\textcolor{red}{observations of SEDs}} and also made classification of galaxies more complex.
Since no two galaxies, even with the same morphology, have exactly the same properties, classifying SEDs of galaxies using templates is very challenging.
To overcome this challenge, many fitting methods have been developed and used to find template matches for SEDs, with $\chi^2$ minimization being the most commonly used. 
Artificial neural networks, K-means clustering, and principal component analysis are some other methods used to cluster and classify spectral types of galaxies \citep[e.g.][]{Allen13,Ordov14,Shi15}.

@@

%SOM
A Kohonen self-organizing map (also called self-organizing map, or SOM) is an \textbf{(semi)-}unsupervised neural network \textbf{\textcolor{red}{for mapping and visualizing a complex and nonlinear high dimension data introduced by~\citet{Kohonen82}.}}
It shows simple geometrical relationships in non-linear high dimensional data on a map \citep{Kohonen98}.
\textbf{An SOM can be trained unsupervised, and its results can be used to classify other data.
Therefore, this method can be considered as (semi)-unsupervised.}

@@

%SOM in Astronomy continued.
Large spectroscopic surveys have made available integrated spectra of millions of galaxies.
These integrated spectra combine the light of billions of individual stars and nebulae within a galaxy, and
finding patterns and common characteristics between galaxies can be a complex task.
\citet{In12} introduced a new clustering tool based on the SOM method for analyzing these large datasets.
They used $\sim 60000$ spectra from the Sloan Digital Sky Survey \citep[SDSS;][]{Abazajian09} to test their tool, and created very large SOMs to analyze the type of spectra/objects.
They also generated SOMs from quasars' spectra in order to find unusual types of spectra. 
Later, \citet{Meusinger16} used these SOMs and updated data from SDSS and other surveys to find a new class of quasars.
The other application of SOMs is to find outliers or errors in the data.
\citet{Fustes13} produced a package based on SOM to classify spectra from the GAIA survey that were previously classified as ``unknown'' by the SDSS pipeline. This package can \textbf{\textcolor{red}{distinguish an astronomical object from instrumental errors,}} and then classify the object based on its spectrum.

%What \citetalias{Hossein12} did
~\citet[][hereafter \citetalias{Hossein12}]{Hossein12} classified \textbf{\textcolor{red}{spectral types}} of a sample of 142 galaxies using a supervised (Multi-layer feed-forward) neural network method, based on the spectral templates presented by \citetalias{Kinney96}.
With the supervised method, they could only classify \textbf{\textcolor{red}{spectral types}} of 105 out of the 142 galaxies.
The \textbf{\textcolor{red}{spectra}} of the 37 galaxies from their samples could not be matched with any of those in the \citetalias{Kinney96} templates. 
In order to classify the remaining 37 galaxies, they combined spectra of the \citetalias{Kinney96} templates.
\citetalias{Hossein12} also showed that there are tight correlations between physical properties of galaxies, and these correlations might be different for each type of galaxy.

%What we did
In this paper, we compare supervised and unsupervised methods of galaxy spectral classification directly, using exactly the same data as \citetalias{Hossein12}.  
Our hypothesis was that since a self-organizing map has the freedom to classify objects in between known classes, it could be applied to this data 
to find the best \textbf{\textcolor{red}{spectral class}} for the 37 galaxies that could not be classified by the supervised method.
First, we train self-organizing maps using the \citetalias{Kinney96} spectral templates and compare our classification with the \citetalias{Kinney96} classification.
Then, we use the trained network to classify \textbf{\textcolor{red}{the spectra}} of 142 galaxies with $0.5 < z < 1$ from \citetalias{Hossein12}. 
As in \citetalias{Hossein12}, we examine the averaged properties of the new galaxy groups and compare them with previous work.
In Section $\S$~\ref{sec: data_highZ}, we present the data that we used to train our networks and the data classified using the trained networks. 
We describe the SOM methods in Section $\S$~\ref{sec: method_somz}. 
The results of the \textbf{\textcolor{red}{spectral classifications}} and a comparison with previous studies are presented in Section $\S$~\ref{sec: result}. 
In Section $\S$~\ref{sec: summary_SOMZ}, we summarize our results and discuss potential future work in this subject.

>>>>

%%Creating templates and NIR_UV templates
Several attempts have been made to create a complete set of templates for categorizing the spectral type of galaxies using \textbf{\textcolor{red}{observations of}} nearby galaxies~(e.g.~\citealt{Kinney93}, \citealt[][hereafter \citetalias{Kinney96}]{Kinney96}, \citealt{Bershady00}, \citealt{Mannucci01}). 
Based on their usage, these templates are restricted to certain wavelengths.
In the near-infrared (NIR) to ultraviolet (UV) wavelengths \textbf{\textcolor{red}{(the region where the energy output of stars peaks)}}, the \textbf{\textcolor{red}{spectrum}}\footnote{\textbf{\textcolor{red}{In the remainder of this paper, we are concerned with a limited range in wavelength, and so we refer to spectra rather than spectral energy distributions.}}} contains information about the main physical properties of galaxies, e.g., age, star formation rate (SFR), stellar mass, metallicity, and interstellar medium.

@@

Advances in techniques and detectors have resulted in more detailed \textbf{\textcolor{red}{observations}} and also made classification of galaxies more complex.
Since no two galaxies, even with the same morphology, have exactly the same properties, classifying \textbf{\textcolor{red}{spectra}} of galaxies using templates is very challenging.
To overcome this challenge, many fitting methods have been developed and used to find template matches for \textbf{\textcolor{red}{spectra}}, with $\chi^2$ minimization being the most commonly used. 
Artificial neural networks, K-means clustering, and principal component analysis are some other methods used to cluster and classify spectral types of galaxies \citep[e.g.][]{Allen13,Ordov14,Shi15}.

@@

%SOM
A Kohonen self-organizing map (also called self-organizing map, or SOM) is an \textbf{(semi)-}supervised neural network \textbf{\textcolor{red}{for mapping and visualizing a complex and nonlinear high dimension data introduced by~\citet{Kohonen82}.}}
It shows simple geometrical relationships in non-linear high dimensional data on a map \citep{Kohonen98}.
\textbf{An SOM can be trained unsupervised, and its results can be used to classify other data.
Therefore, this method can be considered as (semi)-supervised.}

@@

%SOM in Astronomy continued.
Large spectroscopic surveys have made available integrated spectra of millions of galaxies.
These integrated spectra combine the light of billions of individual stars and nebulae within a galaxy, and
finding patterns and common characteristics between galaxies can be a complex task.
\citet{In12} introduced a new clustering tool based on the SOM method for analyzing these large datasets.
They used $\sim 60000$ spectra from the Sloan Digital Sky Survey \citep[SDSS;][]{Abazajian09} to test their tool, and created very large SOMs to analyze the type of spectra/objects.
They also generated SOMs from quasars' spectra in order to find unusual types of spectra. 
Later, \citet{Meusinger16} used these SOMs and updated data from SDSS and other surveys to find a new class of quasars.
The other application of SOMs is to find outliers or errors in the data.
\citet{Fustes13} produced a package based on SOM to classify spectra from the GAIA survey that were previously classified as ``unknown'' by the SDSS pipeline. This package can \textbf{\textcolor{red}{distinguish an astronomical object from instrumental artifact,}} and then classify the object based on its spectrum.

%K-means vs. SOM
\textbf{K-means clustering is another unsupervised method that is being used in many astronomical studies \citep[e.g.][]{DAbrusco12,Ordov14,Boersma14,Aycha16}.
K-mean clustering divides data into K clusters, in the way that each data belong to the cluster with the nearest mean value.
In the first order, both SOM and K-means work similarly.
In fact, SOM is a combination of K-means and considering the effect of the neighbors which can be ``controlled" by the neighborhood functions (i.e., smoothing process). 
In SOM approach a neighborhood function has an important role in the operation of the SOM. 
When the radius of the neighbourhood function goes to zero, the algorithm will lose its ordering power and will act as k-means clustering.
Because of the smoothing process, the SOM can outperform the K-means clustering even in a simple 1D SOM, which can provide a big picture of clustering similar to the result of K-means clustering.}

%What \citetalias{Hossein12} did
~\citet[][hereafter \citetalias{Hossein12}]{Hossein12} classified \textbf{\textcolor{red}{spectral types}} of a sample of 142 galaxies using a supervised (multi-layer feed-forward) neural network method, based on the spectral templates presented by \citetalias{Kinney96}.
The supervised method could only classify \textbf{\textcolor{red}{spectral types}} of 105 out of the 142 galaxies.
The \textbf{\textcolor{red}{spectra}} of the 37 galaxies from their samples could not be matched with any of those in the \citetalias{Kinney96} templates. 
In order to classify the remaining 37 galaxies, they combined spectra of the \citetalias{Kinney96} templates.
\citetalias{Hossein12} also showed that there are tight correlations between physical properties of galaxies, and these correlations might be different for each type of galaxy.

%What we did
In this paper, we compare supervised and unsupervised methods of galaxy spectral classification directly, using exactly the same data as \citetalias{Hossein12}.  
Our hypothesis was that since a self-organizing map has the freedom to classify objects in between known classes, it could be applied to this data 
to find the best \textbf{\textcolor{red}{spectral class}} for the 37 galaxies that could not be classified by the supervised method.
First, we train self-organizing maps using the \citetalias{Kinney96} spectral templates and compare our classification with the \citetalias{Kinney96} classification.
Then, we use the trained network to classify \textbf{\textcolor{red}{the spectra}} of 142 galaxies with $0.5 < z < 1$ from \citetalias{Hossein12}. 
As in \citetalias{Hossein12}, we examine the averaged properties of the new galaxy groups and compare them with previous work.
\textbf{We cluster both \citetalias{Kinney96} templates and \citetalias{Hossein12} data, usign K-means clustering method and compare the results with the results from SOM clustering.}
In Section~\ref{sec: data_highZ}, we present the data that we used to train our networks and the data classified using the trained networks. 
We describe the SOM and \textbf{K-means clustering methods} in Section~\ref{sec: method_somz}. 
The results of the \textbf{\textcolor{red}{spectral classifications}} and a comparison with previous studies are presented in Section~\ref{sec: result}. 
In Section~\ref{sec: summary_SOMZ}, we summarize our results and discuss potential future work in this subject.

### 2016-10-26 sections0.07/method0.07.tex
Final corrections after referee report.

 The self-organizing map is a clustering method which reduces the dimensionality of the data, usually to one or two dimensions (1D or 2D), while preserving topological features of the original dataset.
 The result of an SOM is a set of nodes (neurons) that are arranged in a 1D or 2D arrays \citep{Kohonen98}. 
 Each node may contain one or more samples from the input data.
 The distance between the nodes represents similarity or dissimilarity of the underlying samples, i.e., similar data are closer together in the array and the distance between two nodes is related to the dissimilarity of their samples.
 A weight vector ``\boldit{W}" with the same dimension as the input data is associated with each node and will be varied during the training process.
 This vector is the key factor in determining the position of the nodes in a map.
 \cite{Geach12} presented the application of the self-organized map and discussed its algorithm in detail.
 In this section we briefly discuss the algorithm of SOM, how we create our maps.

@@

     We start by creating S1 $\times$ S2 empty neurons. 
     The arrangement of these neurons depends on the map's topology provided by the user. 
     In the case of 1D maps, since each neuron has two immediate neighbours, the topology of the map does not have any effect on the final result and any topology can be chosen.
     However, in 2D maps, the shape of the neurons specifies the number of immediate neighbours for each neuron and it is up to user to choose the most suitable shape based on the data.
     In this paper, we choose hexagonal topology, which gives each neuron six neighbours, and provides more interactions between neurons.
     Initially a random weight vector, \boldit{W} $\in \Re^n$, will be assigned to each node.
     The process of creating SOM happens over a series of $N$ iterations, where $N$ is set by the user. 
     During each iteration the weight vectors might change according to the Kohonen learning rule (equation~\ref{equ: weight adj}). 
      In each iteration, the SOM code:
     \begin{enumerate}
        \item chooses a random vector from the dataset ($V_i$).
        \item calculates the Euclidean distance in $\in \Re^n$ space for each node, $j$, as  $D_j^2= \sum_{i=0}^{i=n} (V_i - W_i)^2$, and finds a neuron with minimum $D_j$, (``$D_{j_{min}}$"). This neuron is the winner node and is called the Best Matching Unit (BMU). 
        \item  computes the radius of the neighbourhood of the BMU to find nodes within this radius. The weight vectors of these nodes will be affected in the next steps. The radius of the neighbourhood is arbitrary and can be set to be as high as half of the SOM size. It then decays exponentially over each iteration as
        \begin{equation}
            r^t_{BMU} = r^0_{BMU}e^{(-t/\tau)}
        \end{equation}
        where $\tau$ is a decay constant and is usually set to be the same as the number of iterations, $N$. $r^0_{BMU}$ and $r^t_{BMU}$ are the radii of the neighbourhood at 0th and $t$th iteration, respectively. 
        \item changes the weight vectors of the BMU and all the nodes within $r^t_{BMU}$ as:
        \begin{equation}
            \label{equ: weight adj}
            w(t+1)=w(t)+L(t) \times R(t) \times(v(t)-w(t))
        \end{equation}
        where $L(t) = L_0 e^{(-t/\tau)}$ is the learning factor, which prevents the divergence of the SOM and $R(t)=\exp(-\frac{D_j^2}{2r^t_{BMU}})$ is the influence rate. $R(t)$ determines how the weight of each node in the neighbourhood of BMU will change.
     \end{enumerate}
     These steps are then repeated $N$ times.

\subsection{Creating self-organizing maps}
\label{sec: create_som}
     In order to create SOMs, we use the {\sc matlab} neural network toolbox~\citep[NNT,][]{matlabtolbox}. 
     An SOM in {\sc nnt} can be created by the {\sc newsom} or {\sc selforgmap} libraries, both of which work in two phases, an ``ordering phase" and a ``tuning phase". 
     The first phase is called the ``ordering phase" and
     starts with maximum neighbourhood distance and an initial high learning factor (usually 0.9) is provided by the user. 
     The ordering phase continues for a requested number of iterations.
     During the iterations, the learning factor decreases to the tuning phase learning factor and the neighbourhood distance reaches that of the tuning phase as well.
     Both the learning factor and the neighbourhood of the tuning phase are set by the user. 
     The amount by which these two factors change in each iteration depends on the number of iterations.

@@

     We create the final SOMs with initial values for number of iterations in ordering phase, ordering phase learning factor, tuning phase learning factor, and tuning phase neighbourhood distance of 1000, 0.9, 0.02, and 1, respectively. 
     To present our results, we use {\sc nnt}'s built-in plotting tool.
     Specifically, we use two of the plots in this tool: a hits map, which shows the number of times each neuron has became the winner (hits), and a distance map, which shows the distance between those neurons.
     In the maps, the purple hexagonal shapes represent the neurons. 
     The distances in a distance map are shown by the grey cycle colours:
     the darker the colour, the larger the distance between neurons.
     In the hit maps, neurons with zero hits are left empty.

>>>>

 The self-organizing map is a clustering method which reduces the dimensionality of the data, usually to one or two dimensions (1D or 2D), while preserving topological features of the original dataset.
 The result of an SOM is a set of nodes (neurons) that are arranged in a 1D or 2D arrays \citep{Kohonen98}. 
 Each node may contain one or more samples from the input data.
 The distance between the nodes represents similarity or dissimilarity of the underlying samples, i.e., similar data are closer together in the array and the distance between two nodes is related to the dissimilarity of their samples.
 A weight vector ``\boldit{W}" with the same dimension as the input data is associated with each node and will be varied during the training process.
 This vector is the key factor in determining the position of the nodes in a map.
 \cite{Geach12} presented the application of the self-organized map and discussed its algorithm in detail.
 In this section we briefly discuss the algorithm of SOM and how we create our maps.

@@

     We start by creating S1 $\times$ S2 empty neurons. 
     The arrangement of these neurons depends on the map's topology provided by the user. 
     In the case of 1D maps, since each neuron has two immediate neighbours, the topology of the map does not have any effect on the final result and any topology can be chosen.
     However, in 2D maps, the shape of the neurons specifies the number of immediate neighbours for each neuron and it is up to user to choose the most suitable shape based on the data.
     In this paper, we choose hexagonal topology, which gives each neuron six neighbours, and provides more interactions between neurons.
     Initially a random weight vector, \boldit{W} $\in \Re^n$, will be assigned to each node.
     The process of creating SOM happens over a series of $N$ iterations, where $N$ is set by the user. 
     During each iteration the weight vectors might change according to the Kohonen learning rule (equation~\ref{equ: weight adj}). 
      In each iteration, the SOM code:
     \begin{enumerate}
        \item chooses a random vector from the dataset ($V_i$).
        \item calculates the Euclidean distance in $\in \Re^n$ space for each node, $j$, as  $D_j^2= \sum_{i=0}^{i=n} (V_i - W_{j_i})^2$, and finds a neuron with minimum $D_j$, (``$D_{j_{min}}$"). This neuron is the winner node and is called the Best Matching Unit (BMU). 
        \item  computes the radius of the neighbourhood of the BMU to find nodes within this radius. The weight vectors of these nodes will be affected in the next steps. The radius of the neighbourhood is arbitrary and can be set to be as high as half of the SOM size. It then decays exponentially over each iteration as
        \begin{equation}
            r^t_{BMU} = r^0_{BMU}e^{(-t/\tau)}
        \end{equation}
        where $\tau$ is a decay constant and is usually set to be the same as the number of iterations, $N$. $r^0_{BMU}$ and $r^t_{BMU}$ are the radii of the neighbourhood at 0th and $t$th iteration, respectively. 
        \item changes the weight vectors of the BMU and all the nodes within $r^t_{BMU}$ as:
        \begin{equation}
            \label{equ: weight adj}
            w(t+1)=w(t)+L(t) \times R(t) \times(v(t)-w(t))
        \end{equation}
        where $L(t) = L_0 e^{(-t/\tau)}$ is the learning factor, which prevents the divergence of the SOM and $R(t)=\exp(-\frac{D_j^2}{2r^t_{BMU}})$ is the influence rate. $R(t)$ determines how the weight of each node in the neighbourhood of BMU will change.
     \end{enumerate}
     These steps are then repeated $N$ times.

\subsection{Creating self-organizing maps}
\label{sec: create_som}
     In order to create SOMs, we use the {\sc matlab} neural network toolbox~\citep[NNT,][]{matlabtolbox}, written for \textsc{matlab2015b}. 
     An SOM in {\sc nnt} can be created by the {\sc newsom} or {\sc selforgmap} libraries, both of which work in two phases, an ``ordering phase" and a ``tuning phase". 
     The first phase is called the ``ordering phase" and
     starts with maximum neighbourhood distance and an initial high learning factor (usually 0.9) is provided by the user. 
     The ordering phase continues for a requested number of iterations.
     During the iterations, the learning factor decreases to the tuning phase learning factor and the neighbourhood distance reaches that of the tuning phase as well.
     Both the learning factor and the neighbourhood of the tuning phase are set by the user. 
     The amount by which these two factors change in each iteration depends on the number of iterations.

@@

     We create the final SOMs with initial values for number of iterations in ordering phase, ordering phase learning factor, tuning phase learning factor, and tuning phase neighbourhood distance of 1000, 0.9, 0.02, and 1, respectively. 
     To present our results, we use {\sc nnt}'s built-in plotting tool.
     Specifically, we use two of the plots in this tool: a hits map, which shows the number of times each neuron has became the winner (hits), and a distance map, which shows the distance between those neurons.
     In the maps, the coloured hexagonal shapes represent the neurons. 
     The distances in a distance map are shown by the grey cycle colours:
     the darker the colour, the larger the distance between neurons.
     In the hit maps, neurons with zero hits are left empty.

\subsection{K-means clustering}
\label{sec: kmeans_method}     
\textbf{In the K-means method, the user decides in advance on a number of clusters, $K$, which cannot be greater than the dimension of data. 
The algorithm randomly chooses $K$ random points to be set as the centroids of the clusters.
Then it finds the data points that are closest to the each assumed centroid and clusters them together.
Mean values of the data points assigned to each cluster are calculated and become the new centroid of the cluster. 
The algorithm finds new clusters based on the new centroids and repeats the procedure until it converges.
We used \textsc{matlab} K-means library~\citep{Seber84, Spath85}, written for \textsc{matlab2015b}, to perform the K-means clustering.}
